Solving \\ the \\ 15-Puzzle \\ \& \\ Rubik's Cube

### 2022-09-12 latex/Chapters/SlidingPuzzle.tex
update
trimming SP

In this chapter, I show and discuss results on \textbf{SP} of various dimensions, going from \textit{small} dimensions (which I define as those I can reasonably \textit{fully} solve given my computing resources, the finite time I have to complete this project, and the choice of programming language), then I move to studying in much detail the case of the 3x3, and finally move on to slightly higher dimensions 3x4 and 4x4. Table \ref{tab:gridSP} below shows which solvers I have run on each of the dimensions attempted. The reader is referred to the implementation chapter \ref{sec:Implementation} for a detailed explanation of each of these solvers, and to the appendix section \ref{sec:Examples} for code examples on how to run them.

@@

\noindent I am neither making claims of depth (each solver could surely be tuned and optimized/adapted, possibly differently for each dimension) nor breadth (I have not run all of the solvers versus each dimension). Running these experiments takes a lot of computing time (even though I have very decent computing resources), so I had to be somewhat selective. I wanted however to run all of the solvers I have implemented one one particular dimension (the intermediary 3x3 seemed appropriate to do so) and try to answer as many of the following interesting questions as possible:
\begin{itemize}
\item How hard is too hard for \textbf{BFS} to complete in \textit{reasonable} time?
\item How much, if any, improvement do we get with Manhattan++ over Manhattan?
\item Is there much loss between \textbf{DL} and \textbf{DRL}, i.e. is losing the teacher used by \textbf{DL} a deal breaker?
\item Is \textbf{DQL}, all things being equal, able to learn a better cost-to-go heuristic than \textbf{DRL}?
\item How does \textbf{MCTS} perform? How important is it to tune the hyper-parameter $c$? How much improvement, if any, does the post trim of the \textbf{MCTS} tree via \textbf{BFS} bring?
\item Are any of the \textbf{D*L} techniques able to perform on par with A$^{*}$ with Manhattan++ or perfect heuristic?
\end{itemize}
All of these questions are answered (obviously not in generality, but in the context of the experiments I have been running) in this chapter.

@@

As mentioned in chapter \ref{sec:Puzzles}, the state space cardinality for the \textbf{SP} grows very quickly with n and m. The only dimensions which have less than 239.5 millions states are shown in table \ref{tab:smallSP}. Note I am also only considering n $\leq$ m since dimension mxn can always be solved by symmetry from the solutions mxn:

\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 12 & 360 & 20,160 & 1,814,400 \\
3              &   &   & 181,440 &  &    \\
\end{tabular}
\caption{\label{tab:smallSP}\# puzzles for \textit{small} dimensions}
\end{center}
\end{table}
\noindent In this section, I will discuss \textit{full} results for these 5 dimensions. In order to fully solve them, one can simply use rubiks.scripts.learner, setting up the PerfectLearner with A* and manhattan heuristic, or instantiate directly a PerfectLearner as have seen in section \ref{PLSS}. I obtained the following God numbers (table \ref{tab:smallSPGN}) for these puzzles:
\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 6 & 21 & 36 & 55* \\
3              &   &   & 31 &  &    \\
* \tiny{provisional result}
\end{tabular}
\caption{\label{tab:smallSPGN}God numbers for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Among the above dimensions, 2x5 has the largest state space. My current result in table \ref{tab:smallSPGN} for that dimension is provisional, as I have only solved about 1.2 million puzzles out of the 1.8 million possible configurations, despite running a combined 600 hours (most of the time over 16 cores, except for a full week-end run on a 72-core c5.18xlarge Amazon EC2 instance). It is well possible that I haven't yet bumped into the hardest configuration though!
\\
\\
The perfectLearner also keeps track of the hardest puzzle it has encountered (i.e. a configuration whose optimal solution has a cost equal to the God number). I obtained the following hardest puzzles for each of the 5 \textit{small} dimensions:
\\
\begin{three}
\centering
\setrow{2}{,3}
\setrow{1}{2,1}
\end{three}
\begin{five}
\setrow{2}{4,5,}
\setrow{1}{1,2,3}
\end{five}
\begin{seven}
\setrow{2}{,7,2,1}
\setrow{1}{4,3,6,5}
\end{seven}
\begin{nine}
\setrow{2}{,9,3,7,1}
\setrow{1}{5,4,8,2,6}
\end{nine}
\begin{eight}
\setrow{3}{8,6,7}
\setrow{2}{2,5,4}
\setrow{1}{3,,1}
\end{eight}

\Subsection{Manhattan heuristic}
\label{MHComp}
In this section, I verify empirically that, as expected, the overhead of adding penalty in Manhattan++ for the linear constraint (which have all been precomputed and stored in a database) is more than compensated for by the reduction in nodes expansion. I have run my solver script for 2x5 in performance test mode, for both Manhattan and Manhattan++, with 250 randomly shuffled puzzles with \textit{nb\_shuffles} from 0 to 60 by increment of 5, as well as with \textit{nb\_shuffles = $\inf$}. The resulting run time and nodes expansions are as follows:

@@

\noindent As can be seen, for low difficulty (up to \textit{nb\_shuffles = 20}), the node expansions are about the same in both cases, and the overhead of adding the linear constraints penalty increases the run time. However, for any non trivial case, Manhattan++ outperforms considerably (by a factor of about 2.5). Table \ref{tab:mppOutperformance} below summarizes the results for the 250 \textit{perfectly} shuffled configurations:

@@

\begin{figure}
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.45]{./Figures/33SPnets}}
  \caption[33SPnet]{Architecture used for the \textbf{DL} heuristics training on the 3x3 \textbf{SP}}
  \label{fig:33SPnets}
\end{figure}

>>>>

I now discuss results on the \textbf{SP}, going from \textit{small} dimensions (which I can reasonably \textit{fully} solve given my computing resources, the finite time I have to complete this project, and the choice of programming language), then I move to studying in detail the 3x3 \textbf{SP}, and finally move to higher 3x4 and 4x4 dimensions. Table \ref{tab:gridSP} shows which solvers I have run on each dimension. The reader is referred to sections \ref{sec:Chapter1} and \ref{sec:Implementation} for details on each solver, and to the appendix section \ref{sec:Examples} for code examples.

@@

\noindent I am not making claims of depth (each solver could be tuned and optimized much further) nor breadth (I have not run all solvers versus all dimensions). These experiments take a lot of computing time, so I had to be somewhat selective. I wanted however to run all of the solvers on at least one particular dimension (3x3 seemed appropriate) to try and answer, even if not in generality, as many of the following interesting questions:
\begin{itemize}
\item How hard is too hard for \textbf{BFS} to complete in \textit{reasonable} time?
\item How much, if any, improvement do we get with Manhattan++ over Manhattan?
\item Does \textbf{DL} perform much better than \textbf{DRL} due to supervision?
\item Is \textbf{DQL}, all things being equal, able to learn a better value-heuristic than \textbf{DRL}?
\item How does \textbf{MCTS} perform? How important is it to tune the hyper-parameter $c$? How much improvement does the final \textbf{BFS} tree-trimming bring?
\item Are the \textbf{D*L} heuristics competitive versus Manhattan++ or the perfect heuristic?
\end{itemize}

@@

As mentioned in chapter \ref{sec:Puzzles}, the state space cardinality for the \textbf{SP} grows quickly with n and m. The only dimensions which have less than 239.5 millions states are shown in table \ref{tab:smallSP}.

\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 12 & 360 & 20,160 & 1,814,400 \\
3              &   &   & 181,440 &  &    \\
\end{tabular}
\caption{\label{tab:smallSP}\# puzzles for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Using my PerfectLearner with A* and Manhattan heuristic (see \ref{PLSS} for example of how to run this), I obtained the following God numbers (table \ref{tab:smallSPGN}) for these dimensions:
\begin{table}[H]
\begin{center}
\begin{tabular}{l*{6}{c}r}
n              & m & 2 & 3 & 4 & 5\\
\hline
2              &   & 6 & 21 & 36 & 55* \\
3              &   &   & 31 &  &    \\
* \tiny{provisional result}
\end{tabular}
\caption{\label{tab:smallSPGN}God numbers for \textit{small} dimensions}
\end{center}
\end{table}
\noindent Among those dimensions, 2x5 has the largest state space and my result for it is still provisional, as I have only solved about 1.2 out of 1.8 million puzzles at the time of writing, despite running a combined 600 hours (mostly over 16 cores, except for a full week-end run on a 72-core c5.18xlarge Amazon EC2 instance). It is well possible that I haven't yet bumped into the hardest 2x5 configuration! The perfectLearner also keeps track of the hardest puzzle it has encountered (i.e. a configuration whose optimal solution has a cost equal to the God number). I obtained the following hardest puzzles for each of the 5 \textit{small} dimensions:
\\
\begin{three}
\centering
\setrow{2}{,3}
\setrow{1}{2,1}
\end{three}
\begin{five}
\setrow{2}{4,5,}
\setrow{1}{1,2,3}
\end{five}
\begin{seven}
\setrow{2}{,7,2,1}
\setrow{1}{4,3,6,5}
\end{seven}
\begin{nine}
\setrow{2}{,9,3,7,1}
\setrow{1}{5,4,8,2,6}
\end{nine}
\begin{eight}
\setrow{3}{8,6,7}
\setrow{2}{2,5,4}
\setrow{1}{3,,1}
\end{eight}

\Subsection{Manhattan heuristic}
\label{MHComp}
Next, I verify empirically that, as expected, the overhead of adding penalty in Manhattan++ for the linear constraint is compensated for by the reduction in nodes expansion. I have run my solver script for 2x5 in performance test mode, for both Manhattan and Manhattan++, with 250 randomly shuffled puzzles with \textit{nb\_shuffles} from 0 to 60 by increment of 5, as well as with \textit{nb\_shuffles = $\inf$}. The resulting run time and nodes expansions are shown in table \ref{fig:25SPPerformanceManhattan}:

@@

\noindent For low difficulty (\textit{nb\_shuffles} $\leq$ 20), the node expansions are about the same in both cases, and the overhead of adding the linear constraints penalty increases the run time. However, for harder puzzles, Manhattan++ considerably outperforms, by a factor of 2.5+. Table \ref{tab:mppOutperformance} below summarizes the results for the 250 \textit{perfectly} shuffled configurations:

@@

\begin{figure}
  \noindent
  \makebox[\textwidth]{\includegraphics[scale=0.45]{./Figures/33SPnets}}
  \caption[33SPnet]{Architecture I used for the \textbf{DL} heuristics training on the 3x3 \textbf{SP}}
  \label{fig:33SPnets}
\end{figure}
